---
title: "Codebook - Getting and Cleaning Data Course Project"
author: "Me"
date: "2023-12-13"
output: html_document
---

Step 1: Download the data.

The data required for this project have been previously downloaded from provided link and unzipped on drive.
The working folder has been set, but is not shown here for privacy purpose.
The data lies in "UCI HAR Dataset" folder

Step 2: Load 2 libraries needed for this project

```{r message=FALSE, warning=FALSE}
library(data.table)
library(dplyr)
```

Step 3: Read the files into data frames. 

          ##  From Readme file, the dataset includes the following files:
          ## 'features_info.txt'- Shows information about the variables used on the feature vector - needed to understand files structures
          ## 'features.txt'- the complete list of variables of each feature vector - this has to be used when reading test and training files.
          ## 'activity_labels.txt': Links the class labels with their activity name.
          ## 'train/X_train.txt': Training set using names from function used.
          ## 'train/y_train.txt': Training labels using category names.
          ## 'test/X_test.txt': Test set function using names from function used.
          ## 'test/y_test.txt': Test labels using category names.
          ## 'train/subject_train.txt': Each row identifies the subject who performed the activity for each window sample. Its range is from 1 to 30. 

```{r message=FALSE, warning=FALSE}
features <- read.table("data/UCI HAR Dataset/features.txt", col.names = c("n","functions"))
activity_labels <- read.table("data/UCI HAR Dataset/activity_labels.txt", col.names = c("category", "activity"))
x_train <- read.table("data/UCI HAR Dataset/train/X_train.txt", col.names = features$functions)
y_train <- read.table("data/UCI HAR Dataset/train/y_train.txt", col.names = "category")
x_test <- read.table("data/UCI HAR Dataset/test/X_test.txt", col.names = features$functions)
y_test <- read.table("data/UCI HAR Dataset/test/y_test.txt", col.names = "category")
subject_train <- read.table("data/UCI HAR Dataset/train/subject_train.txt", col.names = "subject")
subject_test <- read.table("data/UCI HAR Dataset/test/subject_test.txt", col.names = "subject")
```

```{r echo=TRUE}
str(features)
str(activity_labels)
str(x_train)
str(y_train)
str(x_test)
str(y_test)
str(subject_train)
str(subject_test)
```

Step 4: Verify if there are duplicates subjects in train and test (as a precaution)

```{r message=FALSE, warning=FALSE}
unique_train <- unique(subject_train)
unique_test <- unique(subject_test)
unique <- rbind(unique_test,unique_train)
if (sum(duplicated(unique)) != 0){
  break
}
```

Step 5: Merges the training and the test sets to create one data set

```{r message=TRUE, warning=FALSE, paged.print=TRUE}
## Before merging train and test data set needs to put subject id and category of test with whose file - using cbind
train <- cbind(subject_train,y_train,x_train)
test <- cbind(subject_test,y_test,x_test)

## full data using rbind
data <- rbind(train, test)
```

```{r echo=TRUE}
str(data)
```

Step 6: Create smaller data set and tidy data

```{r message=FALSE, warning=FALSE}
## Extracts only the measurements on the mean and standard deviation for each measurement. 
##data <- data[, grep("subject|category|mean|std", names(data))]
data <- data %>% select(subject, category, contains("mean"), contains("std"))

## Creating factor variable for category
data$category <- as.factor(data$category)
```

```{r echo=TRUE}
str(data)
```

Step 7: Uses descriptive activity names to name the activities in the data set

```{r message=FALSE, warning=FALSE}
data$category <- factor(data$category, labels = activity_labels$activity)

## change variable name from category to activity 
names(data)[2] <- "activity"
```

Step 8: Appropriately labels the data set with descriptive variable names

```{r message=FALSE, warning=FALSE}
## label data with descriptive names 
names(data)<-gsub("Acc", "_Accelerometer", names(data))
names(data)<-gsub("Gyro", "_Gyroscope", names(data))
names(data)<-gsub("Mag", "_Magnitude", names(data))
names(data)<-gsub("angle", "_Angle", names(data))
names(data)<-gsub("gravity", "_Gravity", names(data))
names(data)<-gsub("Jerk", "_Jerk", names(data))

## weird data name
names(data)<-gsub("BodyBody", "Body", names(data))

## Removes dot.
names(data) <- gsub("\\.", "", names(data))

## metrics
names(data)<-gsub("mean", "_Mean_", names(data))
names(data)<-gsub("std", "_Standard_deviation_", names(data))
names(data)<-gsub("Freq", "_Frequency_", names(data))

## Change start of variables names
names(data)<-gsub("^t", "Time_", names(data))
names(data)<-gsub("^f", "Frequency_", names(data))

## Remove leading _
names(data)<-gsub("_$", "", names(data))

## Fix this
names(data)<-gsub("Mean__Frequency", "Mean_Frequency", names(data))

```


```{r echo=TRUE}
str(data)
```


Step 9: Creates a second, independent tidy data set with the average of each variable for each activity and each subject.

```{r message=FALSE, warning=FALSE}
          final_data <- data %>% group_by(subject, activity) %>% summarise_all(mean)
```

Step 10: Check structure of file

```{r echo=TRUE}
str(final_data)
```

Step 11: Write the tidy table into a text file 

```{r message=FALSE, warning=FALSE}
write.table(final_data, "./data/final_data.txt", row.name=FALSE)
```

